\large{Vipassana for Hackers} \\ \Huge{The Proposal} \\ \large\textit{Version 0.1}

### 2024-06-26 the-proposal/the-proposal.tex
reviewing previous methodology written

@@

Meditation research often has difficulty defining and capturing \textit{expertise},
however, and total hours of experience is still confounded by the rate in which those
hours of experience are accumulated (hours per day). We must capture both, as accurately as
possible. Renunciates potentially have both a high hours-per-day rate and extended
(lifelong) duration of experience. The existence of renunciates forces the very
broad category of \textit{expert} meditators into the terrority of 100,000 hours
of practice, even if we limit practice hours to formal, sitting meditation and
estimate twelve (12) hours per day. If we take the much more modest rate of six (6)
hours per day, assume an adult renunciate can learn Vipassana, at the earliest, from
approximately 20 years of
age, and is now of an age when they may participate in an extended study (say, 50-60
years of age), we arrive at roughly 60,000 hours of practice. If we use the oft-cited
``10,000 hours'' measurement for competence in a subject, we might reasonably
describe our rough spectrum of expertise as such:

\begin{itemize}
	\item Beginner: 0-10,000 hours
	\item Intermediate: 10,000-60,000 hours
	\item Expert: 60,000+ hours
\end{itemize}

@@

This is in addition to a large number of other
methodological issues with studies of Vipassana, as described by Alberto Chiesa in
\textit{Vipassana meditation: systematic review of current evidence},
2010. \cite{chiesa2010vipassana} These include the lack of: study replication,
randomized trials, active plus inactive controls, and double/single blinding.

@@

Last, statistical study of Vipassana, at the intermediate level, is currently not
possible. Meditators who are in the process of transitioning from 10,000 hours of
experience to 60,000 hours of experience are not readily available as research
subjects in significant numbers. Because we have not studied such meditators, we are
as of yet not exactly sure what we might be studying when we do study
them. V.S. Ramachandran put this most succinctly:

@@

The line between ``mere object for brain studies'' and ``research tool'' begins to
blur when we consider using meditation for both simultaneously. However, it seems remiss to neglect
gathering whatever objecting EEG/fMRI/MEG data from the long-term subject-researcher
when that data is so readily available.

@@

Because Vipassana meditators are inherently self-selecting, double blinding is not
possible, nor are randomized trials. Single blinding is possible, and should be
employed in future studies where it is feasible. Combination active/inactive controls
are also possible for all \textit{statistical} studies on Vipassana.

The study proposed is an ultralong (lifelong) study of an individual --- myself. As
far as quantitative data gathered from brain scans is concerned, I am effectively
making myself available as a human guinea pig. As far as qualitative data is
concerned, research will take the form of a deep case study. As I will represent
Ramachandran's ``initial sample'', statistical analysis will not be relevant. If the
study proves productive, future studies targeting a specific hypothesis about
Vipassana in a large sample can use learnings from this study in their statistical
structure.

@@

Qualitative research into the specific nature of human consciousness or the broader
nature of consciousness as it applies to any organism or artificial intelligence can
only be performed directly. In this case, a case study performed by an individual
(myself) will explore the question of consciousness and the mind-body problem
directly, through Vipassana meditation, for a prolonged period of time (the remainder
of my life). Many theories and questions currently exist regarding
consciousness, such as \textit{Theory of Mind}, \textit{Theory of Panpsychism},
\textit{Theory of the Entropic Brain},
the \textit{Simulation Hypothesis}, etc. This study will not address any one of the
many theories of consciousness
directly, nor will it absorb any specific hypothesis. The data recovered from
practicing Vipassana meditation directly for prolonged periods of time will almost
certainly overlap with many such ideas but themes and narratives are likely
to be emergent, rather than conforming to an existing hypothesis of consciousness.

@@

The validity of these emergent observations made during the qualitative portion of
the study can be asserted across most axes of validity: prolonged engagement, rich
descriptions, external audits (from more experienced meditators), identification of
researcher bias, peer debriefing (again from more experienced meditators), and
searching for discrepant evidence are all possible. Respondent validation (member
checking), although possible, may not carry much weight in terms of strengthening the
validity of the study, given an initial sample of one. Triangulation will not be
possible with a sample of one. If the study proves productive, in the future
additional researcher-subjects intent on a lifelong Vipassana practice may strengthen
respondent validation and triangulation for parallel studies. Triangulation of
qualitative analysis of consciousness suffers from the paradox that any individual's
consciousness is only directly observable by that individual and conclusions --- even
those validated by triangulation --- will always be in the third-person.

Data will be collected on a daily basis and themes and narratives regularly collected
with the intention of
describing mental phenomena and the evolution of traits as Vipassana practice
progresses. Monthly or yearly schedules will be decided with a
supervisor but an example daily schedule (subject to change before the study begins)
is available in \textit{Appendix 1: Proposed Daily Schedule}.

@@

\section{REFERENCES}
\section{APPENDICES}
\subsection{Appendix 1: Proposed Daily Schedule}

>>>>

There is a dearth of high-quality scientific meditation research.
Due to decades of misunderstanding the subject, it's estimated that
--- of the 50,000+ papers published on the broad topic of ``meditation'' ---
only a few dozen conform to scientific rigour.
\cite{goleman2017altered,semanticscholar2024meditation}

There are exceptions. Notably, Richard Davidson's 2002 study on Minyur Rinpoche and
the subsequent EEG and fMRI studies of serious Tibetan monks.
\cite{https://doi.org/10.1073/pnas.0407401101}
There have also been effective double-blind trials of householders (the laity),
predominantly studying Mindfulness-Based Stress Reduction (MBSR) and
Mindfulness-based Cognitive Therapy (MBCT), which represent two very mild
techniques of meditation.

Sampling from the extremes is a natural consequence of the accessibility of subjects.
Contemplative monks, serious in their meditation practice, are not easily accessed.
Modern householders tend to peak at 2 or 3 hours of meditation per day,
due to other responsibilities.
This is the dichotomic choice in meditation research:
either sporadically studies monks, who can have tremendous meditative capacity
--- or regularly studies householders, who have little to none.

Bridging this monastic / householder divide requires individuals willing to
(a) dedicate themselves to long periods (years or decades) of continuous meditation
practice with a single technique and
(b) submit themselves to continuous scientific inquiry.

@@

Meditation research often has difficulty defining and capturing \textit{expertise},
however, and total hours of experience is still confounded by the rate in which those
hours of experience are accumulated (hours per day). We must capture both, as accurately as
possible. Renunciates potentially have both a high hours-per-day rate and extended
(lifelong) duration of experience. The existence of renunciates forces the very
broad category of \textit{expert} meditators into the terrority of 100,000 hours
of practice, even if we limit practice hours to formal, sitting meditation and
estimate twelve (12) hours per day. If we take the much more modest rate of six (6)
hours per day, assume an adult renunciate can learn Vipassana, at the earliest, from
approximately 20 years of
age, and is now of an age when they may participate in an extended study (say, 40-60
years of age), we arrive at roughly 60,000 hours of practice.
We can reduce this number to 50,000 to match the upper-bound of groups
categorized in studies by Lumma and Brewer.
\cite{https://doi.org/10.1016/j.ijpsycho.2015.04.017,https://doi.org/10.1073/pnas.1112029108}
If we use the oft-cited
``10,000 hours'' measurement for competence in a subject, we might reasonably
describe our rough spectrum of expertise as such:

\begin{itemize}
	\item Beginner: 0-10,000 hours
	\item Intermediate: 10,000-50,000 hours
	\item Expert: 50,000+ hours
\end{itemize}

@@

This is in addition to a large number of other
methodological issues with studies of Vipassana, as described by Alberto Chiesa in
\textit{Vipassana Meditation: Systematic Review of Current Evidence},
2010. \cite{chiesa2010vipassana} These include the lack of: study replication,
randomized trials, active plus inactive controls, and double/single blinding.

@@

Last, statistical study of Vipassana, at the intermediate level, is currently not
possible. Meditators who are in the process of transitioning from 10,000 hours of
experience to 50,000 hours of experience are not readily available as research
subjects in significant numbers. Because we have not studied such meditators, we are
as of yet not exactly sure what we might be studying when we do study
them. V.S. Ramachandran put this most succinctly:

@@

The line between ``mere object for brain studies'' and ``research tool'' begins to
blur when we consider using meditation for both simultaneously. However, we would be
remiss not to
gather whatever objective EEG/fMRI/MEG data from the long-term subject-researcher
when that data is so readily available.

@@

Because Vipassana meditators are inherently self-selecting, double blind trials are not
possible, nor are randomized trials. Single blinding is possible, and should be
employed in future studies where it is feasible. Combination active/inactive controls
are also possible for all \textit{statistical} studies on Vipassana.

The study proposed is an ultralong (lifelong) study of an individual --- myself.
There is potential for others to join this ultralong study, should it show promise.
As far as quantitative data gathered from brain scans is concerned, I am effectively
making myself available as a human guinea pig. As far as qualitative data is
concerned, research will take the form of a deep case study. As I will represent
Ramachandran's ``initial sample'', statistical analysis will not be relevant.
If the study proves productive, future studies targeting a specific hypothesis about
Vipassana in a large sample can use learnings from this study in their statistical
structure.

@@

Qualitative research into the specific nature of human consciousness or the broader
nature of consciousness as it applies to any organism or artificial intelligence can
only be performed directly. In this case, a case study performed by an individual
(myself) will explore the question of consciousness and the mind-body problem
directly, through Vipassana meditation, for a prolonged period of time.
Many theories and questions currently exist regarding
consciousness, such as \textit{Theory of Mind}, \textit{Theory of Panpsychism},
\textit{Theory of the Entropic Brain},
the \textit{Simulation Hypothesis}, etc. This study will not address any one of the
many theories of consciousness
directly, nor will it absorb any specific hypothesis. The data recovered from
practicing Vipassana meditation directly for prolonged periods of time will almost
certainly overlap with many such ideas but themes and narratives are likely
to be emergent, rather than conforming to an existing hypothesis of consciousness.

@@

The validity of these emergent observations made during the qualitative portion of
the study can be asserted across most axes of validity: prolonged engagement, rich
descriptions, external audits (from more experienced meditators), identification of
researcher bias, peer debriefing (again from more experienced meditators), and
searching for discrepancies in evidence are all possible. Respondent validation (member
checking), although possible, may not carry much weight in terms of strengthening the
validity of the study, given an initial sample of one. Triangulation will not be
possible with a sample of one. If the study proves productive, in the future
additional researcher-subjects intent on a lifelong Vipassana practice may strengthen
respondent validation and triangulation for parallel studies. Triangulation of
qualitative analysis of consciousness suffers from the paradox that any individual's
consciousness is only directly observable by that individual and conclusions --- even
those validated by triangulation --- will always be in the third-person.

Data will be collected on a daily basis and themes and narratives regularly collected
with the intention of
describing mental phenomena and the evolution of traits as Vipassana practice
progresses. Monthly or yearly schedules will be decided with a
study supervisor but an example daily schedule
(subject to change) is available in \textit{Appendix 1: Example Daily Schedule}.

@@

\section{REFERENCES}
\section{APPENDICES}

\subsection{Appendix 1: Example Daily Schedule}

EXAMPLE: \\
05:00 - Wake \\
05:30 - 1 hour meditation \\
06:30 - Breakfast \\
07:30 - Exercise \\
09:00 - 3 hours meditation \\
12:00 - Lunch \\
13:00 - 3 hours meditation \\
16:00 - Writing period \\
18:00 - Evening meal \\
19:00 - 1 hour meditation \\
21:00 - Sleep

It is worth noting that this schedule generally persists across weekends and contains
flexibility for longer or shorter writing periods.
It is also worth noting that this is a sample schedule for the beginning of the study.
As the study progresses, it is expected that the duration of sleep will lessen.
